\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-01-20 document/XSGRB_sample.tex
Adding individual bursts

\title{XSGRB - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}}

@@

\abstract{In this work we present the spectra of all $\gamma$-ray bursts (GRBs)
	afterglows that have been observed at the VLT/X-shooter up to \termdate. This
	totals 137 spectroscopic observations of 118 individual bursts, of which
	redshifts have been secured for 91 \% covering a redshift range from 0.059 to
	$\sim$ 8.5. Some of these observation are late-time observations of the
	underlying host galaxy. Based on an set of selection criteria that are unbiased
	with regards to intrinsic properties of the GRBs, the follow-up effort has been
	guided by the \textit{Swift}-telescope to produce a homogeneous sample of
	afterglows for 86 spectra. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric corrections. We provide an
	assessment of the degree of completeness with respect to the parent GRB
	population, in terms of the X-ray properties of the bursts in the sample and
	find that the sample presented here is consistent with being drawn from the
	\textit{Swift} sample. For all bursts (39) for which it is possible, we provide
	a measurement of the hydrogen column density, increasing the total number of
	published HI column density measurements by $\sim$ 50 \%. We constrain the
	fraction of dark bursts to be 27\% - 42\% \todo{Preliminary} and tentatively
	confirm previous results that larger optical darkness is correlated with
	increased X-ray absorption. This dataset provides a unique possibility to study
	the ISM across cosmic time.

}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{RRM observations} \label{RRM}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

For 8 of the observed GRBs in the spectra released here, RRM triggered. One of
the RRM triggers is outside the statistical sample and two of the RRM trigger
are on short GRBs. The fastest response between \swift~GRB trigger and the
beginning of spectroscopic integration is for GRB~160410A, for which the delay
was only 8.4 minutes.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}}
	\caption{Mollweide projection in galactic coordinates of the full sky showing
		the positions on the sky of the bursts presented in this work. The empty
		circles are the positions of all the \textit{Swift} bursts detected until the
		end of 2016. In blue is shown the position of the bursts fulfilling the sample
		criteria specified in Sect. \ref{samplecrit}. The red circles are added around
		GRBs which have been followed up by X-shooter since the commissioning of the
		instrument. Blue points with red circles represent GRBs that enter our sample
		and have X-shooter spectroscopy. The different samples are compared in Sect.
		\ref{results}. The background shows the dust maps presented in
		\citet{Schlegel1998} where the sample criteria cut with $A_V \lesssim 0.5$ mag
		is removed.}
	\label{fig:skymap}
\end{figure*}

@@

The first GRB followed up is GRB090313, observed on the 15th of March, 2009,
during the commissioning of X-shooter UT3. The bursts observed during the
commissioning or science verification (GRB~090313, GRB~090530, GRB~090809,
GRB~090926) are not a part of our statistical sample. The first burst observed
after science verification and X-shooter was moved to UT2, is GRB~091018, which
thereby constitute the first burst entering our statistical sample. For all
bursts that fulfill our sample selection criteria, described in Sect.
\ref{samplecrit}, spectroscopic follow-up has been attempted with X-shooter.
Various conditions can affect our ability to follow up a given burst, and a
discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

For the majority of the bursts, we have observed with a slit width of 1\farc0,
0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively. This sets a
lower limit on the delivered resolution of the spectra based on the tabulated
values of the delivered resolutions, which is 4350, 7450, and 5300 for the UVB,
VIS and NIR-arm
respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}. %The slit width sets the width of the atmospheric sky lines and determines the amount of light lost due to the wavelength-dependent seeing PSF extending outside the coverage of the slit, where the width of sky-lines is always set by the slit width whereas both the delivered resolution and the slitloss changes for the better as the seeing PSF drops below the slit width. For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit, meaning that for almost all observations a slitloss correction is required.
We describe how slitlosses were corrected for in Sect. \ref{postproc}. For
accurate measurements involving line profiles, knowledge of the precise
instrumental resolution is required. The resolution becomes better than the
nominal one, listed above, when the delivered seeing is smaller than the
projected width of the slit on the sky. We discuss how we determine the
effective instrumental resolution in Sect. \ref{resolution}.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
	\caption{Telluric corrected, normalized spectrum of GRB121024A at $z = 2.300$,
	meant to illustrate the typical data quality. The acquisition magnitude is r =
	20, meaning it is in the brighter end of the sample presented here, but not the
	brightest. The spectrum is rich in absorption, including molecular absorption
	of $H_2$. The absorption trough visible at $\sim$ 4000 $\AA$ is due to \lya in
	the host. We have overplot the most prominent lines seen in GRB afterglows from
	\citet{Christensen2011a}. The continuum estimate is shown in dashed red.
	Additionally, three intervening systems are seen in this sightline  This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
	\label{fig:spectrum}
\end{figure*}

@@

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 arcsec$. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to poor signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. While this
simple method is not terribly accurate (for one, the spatial profile of the
trace is not a perfect Gaussian), but it gives a sufficiently accurate estimate
for the analysis of these poor signal to noise science spectra.

@@

Redshift determination of bursts for which the host is too faint for a
spectroscopic redshift measurement relies on the detection of absorption line
imprinted on the GRB afterglow continuum. Because the afterglow rapidly fades
(typically as $\sim t^{-1}$) a rapid follow-up is essential. In Fig.
\ref{fig:timing} we plot the delay from the BAT trigger to the start of the
spectroscopic observation. The shortest delays are observed in RRM-mode. The
fastest follow-up between BAT trigger and start of spectroscopic observations in
our sample is for the short, $z = 1.717$, GRB160410A for which spectroscopic
integration was initialed only 8.4 minutes after the BAT trigger. To illustrate
importance of the follow-up delay for the redshift completeness, we plot the
redshift completeness as a function of delay time in Fig. \ref{fig:timing} for
all the bursts we have followed up, including the ones outside the statistical
sample. Visible is the decreasing fraction of GRBs with a redshift determination
for increasing delays. The overall redshift completeness for bursts that we have
followed up is 92 per cent. This degree of completeness in the followed bursts,
illustrates the efficiency of VLT/X-shooter in redshift determination. Not shown
in the figure are an additional 12 bursts that have redshift determinations
based on late-time host observations with delay times longer than $\sim$10 days.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/NH_dist.pdf}
	\caption{Distributions of DLA hydrogen column densities for DLAs found in
		quasar absorption lines, from \citep{Noterdaeme2012b} in blue. Overplot in green is the kernel density estimate
		is the similar distribution, only for DLAs in GRB sightlines. Values are taken
		from the compilation in Tanvir2017 et al. (in prep)\citet{Tanvir2017}, along
		with the new values presented in this sample. The marginal distributions for
		the two distribution are also shown, where the different environments probed
		are clearly visible in the hydrogen column densities, as previously also noted
		in \citet{Fynbo2009}}
	\label{fig:NH_dist}
\end{figure}

@@

In this paper we have presented the results of a dedicated effort over the
years 2009 -- 2017 to use the X-shooter spectrograph on the ESO-VLT to secure
spectroscopic observations of afterglows and host galaxies of GRBs detected by
{\it Swift}. This work was initiated by a consortium using Guaranteed Time on
X-shooter, but over the years the project continued in open time and the team
was opened to include any researchers interested in contribution to this
effort.

The sample we present here include spectroscopic observations of 86 
systems fulfilling our sample criteria of which 72 are afterglow spectra 
and 14 are observations of the underlying host galaxies. 
UPDATE THIS.
All spectra have been made publicly available in the reduced form used
in this paper.

SWIFT LIMITATIONS
There is some evidence that the burst duration, $T_{90}$, does not scale with
redshift as expected from time-dilation \citep{Kocevski2013, Littlejohns2013a},
suggesting that we could be missing an increasing fraction of the bursts at
higher redshifts, but \citet{Lien2016} argues that we should only be missing the
bright "short" GRBs", but to the degree that \textit{Swift} is complete, we can
test if the sample presented here is complete.

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectographs. The most striking result found is that GRBs
typically are located behind very large column densities of neutral hydrogen.
Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ quasars have a handful such systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified \citep[e.g., five
in][]{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep[][and this work]{Jakobsson2006b,
	Fynbo2009, Cucchiara2015}.  With afterglow spectroscopy (throughout the
electromagnetic spectrum from X-rays to the sub-mm) we hence characterize the
properties of star-forming galaxies over cosmic history in terms of redshifts,
metallicities, molecular contents, ISM temperatures, UV-flux densities,
extinction curves, etc.  A number of independent papers have been published or
submitted for publication focusing on many of these specific issues of our
sample such as extinction curves \citep[][see also \citealt{Fynbo2014,
	Heintz2017}]{Japelj2015, Zafar17}, emission lines from the underlying host
galaxies \citep{Kruhler2015}, the frequency of intervening \ion{Mg}{ii}
absorbers \citep{Christensen2017}, and escape of ionizing radiation
\citep{Tanvir2017}.

It seems important to end this paper by stressing that the potential of 
using GRB sightlines as probes is far from fully harvested. The sample of
sightlines probed by our spectra is not representative for all GRB 
sightlines as we have shown and consistent with earlier findings
from samples based on low-resolution spectroscopy \citep[e.g.,][]{Fynbo2009}
and from studies of complete samples of GRB host galaxies 
\citep{Hjorth2012, Covino2013, Perley2016}. 
\cite{Kruhler2013} convincingly argue, that 
very rich sightlines like that probed by the remarkable GRB\,080607 
\citep{Prochaska2009} are probably significantly more frequent than in the sightlines 
sampled by our spectra. However, with current instrumentation these sightlines are
out of reach except under very fortunate circumstances as in the case of 
GRB\,080607 when the afterglow could be observed only a few minutes after the 
burst with a 10-m class telescope. Observations of such sightlines with 
X-shooter-like spectrographs on th enext generation of 20--40-m telescopes
is likely to be very rewarding.

@@

\subsection{GRB090313 (z = 3.373)}
The first GRB ever observed with X-shooter, during the commissioning of the
instrument, this data formed the basis of GCN
\#9015\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9015.gcn3}} and is published
in \citet{DeUgartePostigo2010}. Due to the lingering brightness of GRB090313,
6.9 ks spectroscopic integration starting 45 hours after the BAT trigger reveals
a wealth of absorption features superposed on the afterglow continuum at a
common redshift of $z = 3.373$. Two intervening systems at $z = 1.959$ and $z =
1.800$ is identified based on strong \mgii-absorption. Because this burst is
observed before the instrument is science-verified, it does not enter into the
statistical sample.

\subsection{GRB090530 (z = 1.266)}
Observed during paranalization of the instrument, this data forms the basis of
GCN \#15571\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/15571.gcn3}}, but is not
published elsewhere. Observations began 20.6 hours after the BAT trigger and 4.8
ks spectroscopic integration in all three arms reveals the absorption signature
for a host at $z = 1.266$ from the detection of \mgii, \mgi, \SIii, \feii,
\aliii. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB090809 (z = 2.737)}
Observed during the first science verification period and was used for GCN
\#9761\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9761.gcn3}} and is
additionally used as the basis for the master thesis by \'Asa Sk\'ulad\'ottir
(2010). 7.2 ks integration starting 10.2 hours after the GRB trigger notice
yields the clear afterglow continuum in all arms from with Lyman Limit located
in the beginning of the UVB coverage. The simultaneous detection of absorption
lines identified as \lya, \SIii, \oi, \SIi$^*$, \SIiv, \civ, \feii, \alii,
\aliii and \mgii at $z = 2.737$ sets it as the redshift of the GRB. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB090926 (z = 2.106)}
Starting during the second science verification period, this dataset forms the
basis of GCN \#9942\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9942.gcn3}} and
is additionally published in \citet{DElia2010}. Spectroscopic integration
started 22 hours after the BAT trigger and from the acquistion camera the
optical afterglow has R = 17.9 mag (vega) at the beginning of the observations
which causes a very continuum to be seen in all arms. An absorption trough due
to \lya is clearly visible along with numerous metal resonance lines \civ,
\SIii, \SIii$^*$ \feii, \mgii, all at $z = 2.106$, marks this as the redshift of
the GRB. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB091018 (z = 0.971)}
The first burst observed during normal operation after science verification was
completed and there is the first burst that enter the statistical sample. This
data is the basis for GCN \#
10042\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10042.gcn3}} and is published
in \citet{Wiersema2012}. With a bright afterglow and a rapid follow-up, this
spectrum is of pristine quality. The afterglow continuum is bright throughout
all spectroscopic arms which allows the ready detection of \alii, \aliii, \feii,
\mnii, \mgii, \mgi, and \caii al located at $z = 0.971$, setting is as the
redshift of the host.

\subsection{GRB091127 (z = 0.490)}
Obtained 4 days after the burst trigger, this data forms the basis for GCN \#
10233\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10233.gcn3}} and is published
in \citet{Vergani2011}. Due to the late follow-up and a nearby moon, the S/N of
the afterglow continuum is poor especially in the UVB arm, why no clear
absorption lines are detected against the afterglow continuum, although see
\citet{Vergani2011} which report a tentative detection of \mgii. Emission lines
from the underlying host is clearly visible with lines from \oii, \hb, \oiii,
and \ha~all at $z = 0.490$. This bursts is additionally associated with
SN2009nz.

\subsection{GRB100205A  (z = na)}
Observed 3 days after the \textit{Swift} trigger. No afterglow or host detected
in 10.8 ks. GRB likely located at high
redshift\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10399.gcn3}}. The spectrum
has not otherwise been published previously.

\subsection{GRB100219A (z = 4.667)}
The data presented here also formed the basis of GCN \#
10441\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10441.gcn3}} and is published
in \citet{Thone2013}. Observations started 12.5 hours after the \textit{Swift}
trigger and has a total exposure time of 4.8 ks. Absorption features, including
those of \lya~and from a multitude of ions are detected against the afterglow
continuum at $z = 4.667$. Additionally, absorption from an intervening system
is found at $z = 2.181$.

>>>>

\title{XSGRB I - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}}

@@

\abstract{In this work we present the spectra of all $\gamma$-ray bursts (GRBs)
	afterglows that have been observed at the VLT/X-shooter up to \termdate. This
	totals 137 spectroscopic observations of 118 individual bursts, of which
	redshifts have been secured for 91 \% covering a redshift range from 0.059 to
	$\sim$ 8.5. Some of these observation are late-time observations of the
	underlying host galaxy. Based on an set of selection criteria that are unbiased
	with regards to intrinsic properties of the GRBs, the follow-up effort has been
	guided by the \textit{Swift}-telescope to produce a homogeneous sample of
	afterglows for 86 spectra. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric corrections. We provide an
	assessment of the degree of completeness with respect to the parent GRB
	population, in terms of the X-ray properties of the bursts in the sample and
	find that the sample presented here is consistent with being drawn from the
	\textit{Swift} sample. For all bursts (39) for which it is possible, we provide
	a measurement of the hydrogen column density, increasing the total number of
	published HI column density measurements by $\sim$ 50 per cent. We constrain the
	fraction of dark bursts to be less than 28 per cent based on the detection of an
	optical afterglow and tentatively confirm previous results that larger optical
	darkness is correlated with increased X-ray absorption. This dataset provides a
	unique possibility to study the ISM across cosmic time. }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{RRM observations} \label{RRM}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

For 8 of the observed GRBs released here, RRM triggered. One of the RRM triggers
is outside the statistical sample and two of the RRM trigger are on short GRBs.
The fastest response between \swift~GRB trigger and the beginning of
spectroscopic integration is for GRB~160410A, for which the delay was only 8.4
minutes.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. The empty
	circles are the positions of all the \textit{Swift} bursts detected until the
	end of 2016. In blue is shown the position of the bursts fulfilling the sample
	criteria specified in Sect. \ref{samplecrit}. The red circles are added around
	GRBs which have been followed up by X-shooter since the commissioning of the
	instrument. Blue points with red circles represent GRBs that enter our sample
	and have X-shooter spectroscopy. The different samples are compared in Sect.
	\ref{results}. The background shows the dust maps presented in
	\citet{Schlegel1998} where the sample criteria cut with $A_V \lesssim 0.5$ mag
	is removed. The colorbar below indicates the value of $A_V$ on the plot. The
	dotted lines indicate intervals of 30$^\circ$ in longitude and latitude.}
\label{fig:skymap}
\end{figure*}

@@

The first GRB followed up is GRB~090313, observed on the 15th of March, 2009,
during the commissioning of X-shooter UT3. The bursts observed during the
commissioning or science verification (GRB~090313, GRB~090530, GRB~090809,
GRB~090926) are not a part of our statistical sample. The first burst observed
after science verification and X-shooter was moved to UT2, is GRB~091018, which
thereby constitute the first burst entering our statistical sample. For all
bursts that fulfill our sample selection criteria, described in Sect.
\ref{samplecrit}, spectroscopic follow-up has been attempted with X-shooter.
Various conditions can affect our ability to follow up a given burst, and a
discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

For the majority of the bursts, we have observed with a slit width of 1\farc0,
0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively. This sets a
lower limit on the delivered resolution of the spectra based on the tabulated
values of the delivered resolutions, which is 4350, 7450, and 5300 for the UVB,
VIS and NIR-arm
respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}. %The slit width sets the width of the atmospheric sky lines and determines the amount of light lost due to the wavelength-dependent seeing PSF extending outside the coverage of the slit, where the width of sky-lines is always set by the slit width whereas both the delivered resolution and the slitloss changes for the better as the seeing PSF drops below the slit width. For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit, meaning that for almost all observations a slitloss correction is required.
%We describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements involving line profiles, knowledge of the precise
instrumental resolution is required. The resolution becomes better than the
nominal one, listed above, when the delivered seeing is smaller than the
projected width of the slit on the sky. We discuss how we determine the
effective instrumental resolution in Sect. \ref{resolution}.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$,
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$r$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption, including absorption
	from molecular $\mathrm{H_2}$. The absorption trough visible at $\sim$ 4000
	$\AA$ is due to \lya~in the host. We have overplot the most prominent lines
	seen in GRB afterglows from \citet{Christensen2011a}. Additionally, three
	intervening systems are seen in this sightline This spectrum has been analyzed
	in detail in \citet{Friis2015}.} \label{fig:spectrum}
\end{figure*}

@@

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to poor signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. While this
simple method is not terribly accurate (for one, the spatial profile of the
trace is not a perfect Gaussian), but it gives a sufficiently accurate estimate
for the analysis of these poor signal to noise science spectra.

@@

Redshift determination of bursts for which the host is too faint for a
spectroscopic redshift measurement relies on the detection of absorption line
imprinted on the GRB afterglow continuum. Because the afterglow rapidly fades
(typically as $\sim t^{-1}$) a rapid follow-up is essential. In Fig.
\ref{fig:timing} we plot the delay from the BAT trigger to the start of the
spectroscopic observation. The shortest delays are observed in RRM-mode. The
fastest follow-up between BAT trigger and start of spectroscopic observations in
our sample is for the short, $z = 1.717$, GRB~160410A for which spectroscopic
integration was initialed only 8.4 minutes after the BAT trigger. To illustrate
importance of the follow-up delay for the redshift completeness, we plot the
redshift completeness as a function of delay time in Fig. \ref{fig:timing} for
all the bursts we have followed up, including the ones outside the statistical
sample. Visible is the decreasing fraction of GRBs with a redshift determination
for increasing delays. The overall redshift completeness for bursts that we have
followed up is 92 per cent. This degree of completeness in the followed bursts,
illustrates the efficiency of VLT/X-shooter in redshift determination. Not shown
in the figure are an additional 12 bursts that have redshift determinations
based on late-time host observations with delay times longer than $\sim$10 days.

@@

\begin{figure}
	\centering \includegraphics[width=9cm]{figures/NH_dist.pdf}
\caption{Distributions of DLA hydrogen column densities for DLAs found in
	quasar absorption lines, from \citep{Noterdaeme2012b} in blue. Overplot in
	green is the kernel density estimate is the similar distribution, only for DLAs
	in GRB sightlines. Values are taken from the compilation in Tanvir2017 et al.
	(in prep), along with the new values presented in this sample. The marginal
	distributions for the two distribution are also shown along the left side and
	on the top, where the different environments probed are clearly visible in the
	hydrogen column densities, as previously also noted in \citet{Fynbo2009}}
\label{fig:NH_dist}
\end{figure}

@@

In this paper we have presented the results of a dedicated effort over the years
2009 -- 2017 to use the X-shooter spectrograph on the ESO-VLT to secure
spectroscopic observations of afterglows and host galaxies of GRBs detected by
\swift. This work was initiated by a consortium using Guaranteed Time on
X-shooter, but over the years the project continued in open time and the team
was opened to include any researchers interested in contribution to this effort.

The sample we present here include spectroscopic observations of 90 systems
fulfilling our sample criteria including 17 spectra that are late time
observations of the underlying host galaxies. All spectra have been made
publicly available in the reduced form used in this paper.

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectographs. The most striking result found is that GRBs
typically are located behind very large column densities of neutral hydrogen.
Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ quasars have a handful such systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified \citep[e.g., five
in][]{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep[][and this work]{Jakobsson2006b,
	Fynbo2009, Cucchiara2015}.  With afterglow spectroscopy (throughout the
electromagnetic spectrum from X-rays to the sub-mm) we hence characterize the
properties of star-forming galaxies over cosmic history in terms of redshifts,
metallicities, molecular contents, ISM temperatures, UV-flux densities,
extinction curves, etc.  A number of independent papers have been published or
submitted for publication focusing on many of these specific issues of our
sample such as extinction curves \citep[][see also \citealt{Fynbo2014,
	Heintz2017}]{Japelj2015, Zafar17}, emission lines from the underlying host
galaxies \citep{Kruhler2015}, the frequency of intervening \mgii~ absorbers
\citep{Christensen2017}, and escape of ionizing radiation (Tanvir et al. in
prep). %\citep{Tanvir2017}.

%SWIFT LIMITATIONS
%There is some evidence that the burst duration, $T_{90}$, does not scale with
%redshift as expected from time-dilation \citep{Kocevski2013, Littlejohns2013a},
%suggesting that we could be missing an increasing fraction of the bursts at
%higher redshifts, but \citet{Lien2016} argues that we should only be missing the
%bright "short" GRBs", but to the degree that \textit{Swift} is complete, we can
%test if the sample presented here is complete.

It seems important to end this paper by stressing that the potential of using
GRB sightlines as probes is far from fully harvested. The sample of sightlines
probed by our spectra is not representative for all GRB sightlines as we have
shown and consistent with earlier findings from samples based on low-resolution
spectroscopy \citep[e.g.,][]{Fynbo2009} and from studies of complete samples of
GRB host galaxies \citep{Hjorth2012, Covino2013, Perley2016}. \cite{Kruhler2013}
convincingly argue, that very rich sightlines like that probed by the remarkable
GRB~080607 \citep{Prochaska2009} are probably significantly more frequent than
in the sightlines sampled by our spectra. However, with current instrumentation
these sightlines are out of reach except under very fortunate circumstances as
in the case of GRB~080607 when the afterglow could be observed only a few
minutes after the burst with a 10-m class telescope. Observations of such
sightlines with X-shooter-like spectrographs on th enext generation of 20--40-m
telescopes is likely to be very rewarding.

@@

\subsection{GRB~090313 (z = 3.373)}
The first GRB ever observed with X-shooter, during the commissioning of the
instrument, this data formed the basis of GCN
\#9015\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9015.gcn3}} and is published
in \citet{DeUgartePostigo2010}. Due to the lingering brightness of GRB~090313,
6.9 ks spectroscopic integration starting 45 hours after the BAT trigger reveals
a wealth of absorption features superposed on the afterglow continuum at a
common redshift of $z = 3.373$. Two intervening systems at $z = 1.959$ and $z =
1.800$ is identified based on strong \mgii-absorption. Because this burst is
observed before the instrument is science-verified, it does not enter into the
statistical sample.

\subsection{GRB~090530 (z = 1.266)}
Observed during paranalization of the instrument, this data forms the basis of
GCN \#15571\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/15571.gcn3}}, but is not
published elsewhere. Observations began 20.6 hours after the BAT trigger and 4.8
ks spectroscopic integration in all three arms reveals the absorption signature
for a host at $z = 1.266$ from the detection of \mgii, \mgi, \SIii, \feii,
\aliii. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB~090809 (z = 2.737)}
Observed during the first science verification period and was used for GCN
\#9761\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9761.gcn3}} and is
additionally used as the basis for the master thesis by \'Asa Sk\'ulad\'ottir
(2010). 7.2 ks integration starting 10.2 hours after the GRB trigger notice
yields the clear afterglow continuum in all arms from with Lyman Limit located
in the beginning of the UVB coverage. The simultaneous detection of absorption
lines identified as \lya, \SIii, \oi, \SIi$^*$, \SIiv, \civ, \feii, \alii,
\aliii and \mgii at $z = 2.737$ sets it as the redshift of the GRB. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB~090926 (z = 2.106)}
Starting during the second science verification period, this dataset forms the
basis of GCN \#9942\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9942.gcn3}} and
is additionally published in \citet{DElia2010}. Spectroscopic integration
started 22 hours after the BAT trigger and from the acquistion camera the
optical afterglow has R = 17.9 mag (vega) at the beginning of the observations
which causes a very continuum to be seen in all arms. An absorption trough due
to \lya is clearly visible along with numerous metal resonance lines \civ,
\SIii, \SIii$^*$ \feii, \mgii, all at $z = 2.106$, marks this as the redshift of
the GRB. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB~091018 (z = 0.971)}
The first burst observed during normal operation after science verification was
completed and there is the first burst that enter the statistical sample. This
data is the basis for GCN \#
10042\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10042.gcn3}} and is published
in \citet{Wiersema2012}. With a bright afterglow and a rapid follow-up, this
spectrum is of pristine quality. The afterglow continuum is bright throughout
all spectroscopic arms which allows the ready detection of \alii, \aliii, \feii,
\mnii, \mgii, \mgi, and \caii al located at $z = 0.971$, setting is as the
redshift of the host.

\subsection{GRB~091127 (z = 0.490)}
Obtained 4 days after the burst trigger, this data forms the basis for GCN \#
10233\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10233.gcn3}} and is published
in \citet{Vergani2011}. Due to the late follow-up and a nearby moon, the S/N of
the afterglow continuum is poor especially in the UVB arm, why no clear
absorption lines are detected against the afterglow continuum, although see
\citet{Vergani2011} which report a tentative detection of \mgii. Emission lines
from the underlying host is clearly visible with lines from \oii, \hb, \oiii,
and \ha~all at $z = 0.490$. This bursts is additionally associated with
SN2009nz.

\subsection{GRB~100205A  (z = na)}
Observed 3 days after the \textit{Swift} trigger. No afterglow or host detected
in 10.8 ks. GRB likely located at high
redshift\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10399.gcn3}}. The spectrum
has not otherwise been published previously.

\subsection{GRB~100219A (z = 4.667)}
The data presented here also formed the basis of GCN \#
10441\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10441.gcn3}} and is published
in \citet{Thone2013}. Observations started 12.5 hours after the \textit{Swift}
trigger and has a total exposure time of 4.8 ks. Absorption features, including
those of \lya~and from a multitude of ions are detected against the afterglow
continuum at $z = 4.667$. Additionally, absorption from an intervening system
is found at $z = 2.181$.
